Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-19 paper.tex
...

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output,
          native,@const,@pure,@safe,define},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{pingus}
\caption{Pingus gameplay.
\label{fig.sweeney}
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{sweeney}
\caption{Three kinds of code~\cite{games.sweeney}.
\label{fig.sweeney}
}
\end{figure}

@@

Contributions:
    - patterns
    - solutions

@@

We selected 9 representative game behaviors and describe their
implementations in C++ and \CEU.
We also categorized these examples in 5 abstract C++ control-flow patterns that
likely apply to other games:

@@

\subsection{Finite State Machines \\ Case Study: The Armageddon Button Double Click}
%\label{sec.patterns.fsm}

State machines describe the behavior of entities by mapping event occurrences
to transitions between states that trigger appropriate actions.

%\subsection{The Armageddon Button Double Click}
%\label{sec.patterns.fsm.armageddon}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{double-click-opt}
\caption{Double click detection.
\label{fig.armageddon}
}
\end{figure}

In Pingus, a double click in the \emph{Armageddon} button at the bottom right
of the screen literally explodes all pingus (Figure~\ref{fig.armageddon}).
Figure~\ref{fig.armageddon.code} compares the implementations in C++ and \CEU.

\begin{figure*}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
ArmageddonButton::ArmageddonButton(<...>):
    RectComponent(<...>),
    pressed(false); // button initially not pressed
    press_time(0);    // how long since 1st click?
    <...>
{
    <...>
}

@@

void ArmageddonButton::on_click (<...>) {
    if (pressed) {
        server->send_armageddon_event();
    } else {
        pressed = true;
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
do
    var& RectComponent c = <...>;
    <...>
    loop do
        await c.component.on_click;
        watching 1s do
            await c.component.on_click;
            break;
        end
    end
    <...>
    emit outer.game.go_armageddon;
end

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{double-click}
\caption{State machine for the \emph{Armageddon} double click.
\label{fig.armageddon.fsm}
}
\end{figure}

@@

\subsection{Continuation Passing \\ Case Study: Advancing Pages in the Story Screen}

The completion of a long-lasting activity may carry a continuation, i.e., some
action to execute next.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{story-anim}
\caption{The Story screen.
\label{fig.story}
}
\end{figure}

@@

\begin{figure*}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryScreenComponent::StoryScreenComponent (<...>) :
    <...>
{
    pages        = <...>; // vector with loaded pages
    current_page = pages.back(); // first loaded page
    displayed    = false; // if current is complete
    <...>
}

@@

void StoryScreenComponent::update (<...>) {
    <...>
    if (&lt;all-words-appearing&gt;) {
        displayed = true;
    }
}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{story}
\caption{State machine for the Story screen.
\label{fig.story}
}
\end{figure}

@@

\subsection{Dispatching Hierarchies \\ Case Study: Bomber `draw` and `update` callbacks}

Entities typically form a dispatching hierarchy in which a container that
receives a stimulus automatically forwards it to its managed children.

\begin{figure*}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
class Bomber : public PinguAction {
    <...>
    Sprite sprite;
}

@@

void Bomber::draw (SceneContext& gc) {
    <...>
    gc.color().draw(sprite, <...>);
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName do
    <...>
    var&? Sprite sprite = spawn Sprite(<...>);
    <...>
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{story}
\caption{State machine for the Story screen.
\label{fig.story}
}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{hierarchy}
\caption{Dispatching chain for \code{update}.
\label{fig.hier}
}
\end{figure}

@@

\subsection{Lifespan Hierarchies \\ Case Study: The Pingus Container}

Entities typically form a lifespan hierarchy in which a terminating
container entity automatically destroys its managed children.

\begin{figure*}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Pingu* PinguHolder::create_pingu (<...>) {
    <...>
    Pingu* pingu = new Pingu (<...>);
    pingus.push_back(pingu);
    <...>
}

void PinguHolder::update() {
    <...>
    while(pingu != pingus.end()) {
        (*pingu)->update();
        if ((*pingu)->get_status() == Pingu::PS_DEAD) {
            pingu = pingus.erase(pingu);
        }
        <...>
        ++pingu;
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.40\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Game (void) do
    <...>
    pool[] Pingu pingus;
    code/await Pingu_Spawn (<...>) do
        <...>
        spawn Pingu(<...>) in outer.pingus;
    end
    <...>   // code invoking Pingu_Spawn
end

TODO

code/await Pingu (<...>) do
    <...>
    loop do
        await outer.game.dt;
        if call Pingu_Rel_Getpixel(0,-1) == {Groundtype::GP_OUTOFSCREEN} then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{story}
\caption{State machine for the Story screen.
\label{fig.story}
}
\end{figure}

In C++, the class \code{PinguHolder} is a container that holds all pingus
alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{TODO} and also occurs
in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{pool}
\caption{Lifespan of dynamic instances.
\label{fig.pool}
}
\end{figure}

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen spawns a new \code{Pingu} on every invocation of
\code{Pingu\_Spawn}.
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its *natural termination*, the instance is automatically removed
from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. TODO) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

\subsection{Signaling Mechanisms\\ Case Study: Global Keys and the Options Menu}

Entities often need to communicate explicitly through signaling mechanisms,
especially if there is no hierarchy relationship between them.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{options-anim-opt}
\caption{The \emph{Mouse Grab} configuration option.
\label{fig.options}
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{options-anim-opt}
\caption{Mutual dependency between signals.
\label{fig.events}
}
\end{figure}

@@

\begin{figure*}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
void GlobalEvent::on_button_press (<...>) {
    <...>
    switch (event.keysym.sym) {
        case SDLK_g:
            if (keystate[SDLK_LCTRL] || keystate[SDLK_RCTRL]) {
                config_manager.set_mouse_grab(          @ctrl_g_cpp
                    !config_manager.get_mouse_grab());
            }
            break;
        <...>
    }
}

@@

boost::signals2::signal<void(bool)> on_mouse_grab_change;   // definition in `config_manager.h` @signal_def

void ConfigManager::set_mouse_grab (bool v) {   @set_mouse_grab
    <...>
    if (v != get_mouse_grab()) {                @if_1
        <...>   // the actual "grab" effect
        on_mouse_grab_change(v);                @signal
    }                                           @if_2
}

@@

boost::signals2::signal<void (bool)> on_change;   // definition in `check_box.hpp`

>>>>

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,code,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output,
          native,@const,@pure,@safe,define},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus}
\caption{Pingus gameplay.
\label{fig.sweeney}
}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{sweeney}
\caption{Three kinds of code~\cite{games.sweeney}.
\label{fig.sweeney}
}
\end{figure}

@@

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
In Pingus, the game logic also accounts for almost half the size of the
codebase (18.173 from 39.362 LoCs, or 46\%).

rewriting Pingus to \CEU is to

Contributions:
    - present SSRP as an effective alternative for game logic dev
    - qualitative analysis
        - patterns
            - solutions

\section{The Pingus Codebase}

In Pingus, the game logic also accounts for almost half the size of the
codebase (18.173 from 39.362 LoCs, or 46\%).

We rewrote 126 of the 272 files (46\%) which account for 9.186 of the 18.173
LoCs (51\%) comprising the game logic of Pingus [[![X]][cpp\_uncompressed]].
Half of the game logic relates to non-reactive code, such as configurations and
options, saved games and serialization, maps and levels descriptions, string
formatting, collision detection, graph algorithms, etc.
This part remains unchanged and relies on the seamless integration between \CEU
and C/C++ to remain usable.
From the 9.186 touched LoCs, we removed all headers, declarations, trivial
getters \& setters, and other innocuous statements, resulting in 70
implementation files with 4.135 dense LoCs originally written in C++
[[![X]][cpp\_compressed]].
We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs
[[![X]][ceu\_compressed]].
Figure~\ref{tab.tree} summarizes the condensed codebase in the two
implementations:

This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
<!--, helping on encapsulation and cohesion.-->
Nonetheless, the lowest ratio numbers above correlate to the parts of the game
logic that we consider more susceptible to structured reactive programming.
For instance, the *Pingu* behavior (*ratio 0.80*) contains complex animations
that are affected by timers, game rules, and user interaction.
In contrast, the *Option* screen (*ratio 0.97*) is a simple UI grid with
trivial mouse interactions.

\begin{figure*}[t]
\begin{verbatim}
Path              Ceu      C++   Ceu/C++    Descritpion
------------     ----     ----     ----     --------------------------------------------
game/            2064     2268     0.91     the main gameplay
  ./              710      679     1.05         main functionality
  objs/           470      478     0.98         world objects (tiles, traps, etc)
  pingu/          884     1111     0.80         pingu behaviors
    ./            343      458     0.75             main functionality
    actions/      541      653     0.83             pingu actions (bomber, climber, etc)
worldmap/         468      493     0.95     campaign worldmap
screens/         1109     1328     0.84     menus and screens
    option/       347      357     0.97         option menu
    others/       762      971     0.78         other menus and screens
misc/              56       46     1.22     miscellaneous functionality
                 ----     ----     ----
                 3697     4135     0.89
\end{verbatim}
\caption{The Pingus codebase directory tree.
\label{tab.tree}
}
\end{figure*}

%[cpp_uncompressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP
%[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
%[ceu_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL

@@

We selected 6 representative game behaviors and describe their
implementations in C++ and \CEU.
We also categorized these examples in 5 abstract C++ control-flow patterns that
likely apply to other games:

@@

\subsection{Finite State Machines}

\textbf{Case Study: Detecting double-clicks in the \emph{Armageddon button}}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{double-click-opt}
\caption{Double click detection.
\label{fig.armageddon}
}
\end{figure}

In Pingus, a double click in the \emph{Armageddon button} at the bottom right
of the screen literally explodes all pingus (Figure~\ref{fig.armageddon}).
Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.

\begin{figure*}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
ArmageddonButton::ArmageddonButton(<...>):
    RectComponent(<...>),
    pressed(false); // button initially not pressed
    press_time(0);    // how long since 1st click?
    <...>
{
    <...>
}

@@

void ArmageddonButton::on_click (<...>) {
    if (pressed) {
        server->send_armageddon_event();
    } else {
        pressed = true;
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
do
    var& RectComponent c = <...>;
    <...>
    loop do
        await c.component.on_click;
        watching 1s do
            await c.component.on_click;
            break;
        end
    end
    <...>
    emit game.go_armageddon;
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Detecting double-clicks in the \emph{Armageddon button}.
\label{lst.armageddon}
}
\end{figure*}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{double-click}
\caption{State machine for detecting double-clicks in the
         \emph{Armageddon button}.
\label{fig.armageddon.fsm}
}
\end{figure}

@@

\textbf{Case Study: The \emph{Bomber action} sequence}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-anim}
\caption{ The \emph{Bomber action}.
\label{fig.bomber.anim}
}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{bomber-state}
\caption{ State machine for the *Bomber* animation.
\label{fig.bomber.state}
}
\end{figure}

The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius (Figure~\ref{fig.bomber.anim}).
%
We can model the explosion animation with a sequential state machine
(Figure~\ref{fig.bomber.state}) with actions associated to specific frames as
follows:
%
\begin{enumerate}
\item 0th frame:  plays a "Oh no!" sound.
\item 10th frame: plays a "Bomb!" sound.
\item 13th frame: throws particles, destroys the terrain, and shows an
                  explosion sprite.
\item Game tick:  hides the explosion sprite.
\item Last frame: kills the pingu.
\end{enumerate}

TODO
%*([This video][youtube_bomber] presents the sound effects.)*
%[youtube_bomber]: https://youtu.be/QLXIT59il6o?t=306

\begin{figure*}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Bomber::Bomber (Pingu* p) :
    <...>
    sprite(<...>),              // bomber sprite
    sound_played(false),        // tracks state 2               @def_1
    particle_thrown(false),     // tracks state 3
    colmap_exploded(false),     // tracks state 3
    gfx_exploded(false)         // tracks state 4               @def_2
{
    <...>
    // 1. 0th frame: plays a "Oh no!" sound.
    get_world()->play_sound("ohno", pingu->get_pos());          @sound_ohno
}

void Bomber::update () {                                        @update_ini
    sprite.update();                                            @update_1
    <...>   // pingu movement                                   @update_2

    // 2. 10th frame: plays a "Bomb!" sound.                    @sound_bomb_1
    if (sprite.get_current_frame()==10 && !sound_played) {
        sound_played = true;
        get_world()->play_sound("plop", pingu->get_pos());
    }                                                           @sound_bomb_2

    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
    if (sprite.get_current_frame()==13 && !particle_thrown) {   @state_3_1
        particle_thrown = true;
        get_world()->get_pingu_particle_holder()->add_particle(...);
    }
    if (sprite.get_current_frame()==13 && !colmap_exploded) {
        colmap_exploded = true;
        get_world()->remove(bomber_radius, <...>);
    }                                                           @state_3_2

    // 5. Last frame: kills the Pingu
    if (sprite.is_finished ()) {                                @die_1
        pingu->set_status(Pingu::PS_DEAD);
    }                                                           @die_2
}                                                               @update_end

void Bomber::draw (SceneContext& gc) {
    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
    // 4. Game tick: hides the explosion sprite
    if (sprite.get_current_frame()==13 && !gfx_exploded) {      @state_3_3
        gfx_exploded = true;                                    @state_3_4
        gc.color().draw(explo_surf, <...>);                     @state_4
    }
    gc.color().draw(sprite, pingu->get_pos());
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> _ActionName__Enum
do
    <...>
    spawn Mover();                          // spawn the pingu movement to execute in the "background"      @move

    var&? Sprite s = spawn Sprite(<...>);   // spawn the bomber animation to execute in the "background"    @sprite
    watching s do                                                   @watch-1
        // 1. 0th frame: plays a "Oh no!" sound.                    @anim_1
        {Sound::PingusSound::play_sound("ohno", 0.5, 0.0)};

        // 2. 10th frame: plays a "Bomb!" sound.
        await outer.game.dt until s!.sprite.frame == 10;            @frame_1
        {Sound::PingusSound::play_sound("plop", 0.5, 0.0)};

        // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
        await outer.game.dt until s!.sprite.frame == 13;            @frame_2
        spawn PinguParticles(<...>) in outer.pingu_particles;       @particles
        call Game_Remove({&bomber_radius}, <...>);
        do                                                          @do
            <...>
            spawn Sprite(<...>);            // explosion            @explo

            // 4. Game tick: hides the explosion sprite
            await outer.game.dt;                                    @frame_3
        end                                                         @end
        await FOREVER;
    end                                                             @watch-2

    // 5. Last frame: kills the pingu
    escape {ActionName::DEAD};                                      @anim_2
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Advancing pages in the \emph{Story Screen}.
\label{lst.story}
}
\end{figure*}

In C++, the class \code{Bomber} defines the callbacks \code{draw} and
\code{update} to manage the state machine described above:
%
The class first defines one state variable for each action to perform
(ln. 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln. 11).
The \code{update} callback (ln. 14--38) updates the pingu animation and
movement on every frame regardless of its current state (ln. 15--16).
When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
switches to \emph{state-2} (ln. 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln. 25--32 and
43--44).
The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
Finally, the pingu dies after the animation frames terminate (ln. 35--37).
%
Note that a single numeric state variable suffices to track the states, but the
original authors probably chose to encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., *methods + mutable state*).
%
Like the C++ implementation for detecting double clicks, note also that the
state machine is encoded across 3 different methods, each intermixing code with
unrelated functionality.

The equivalent code for the \emph{Bomber action} in \CEU doesn't require state
variables and reflects the sequential state machine implicitly, using
\code{await} statements in direct style to separate the actions.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine~\cite{sync_async.cooperative}: a subroutine that retains runtime
state, such as local variables and the program counter, across reactions to
events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other abstractions
and execute in separate through the \code{spawn} primitive (ln. 4,6).
The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every frame.
%
The code tracks the animation instance (ln. 7--27), performing the last bomber
action on termination (ln. 30).
As soon as the animation starts, the code performs the first action (ln. 8).
The intermediate actions are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all actions happen in a contiguous
chunk of code (ln. 6--30) which handles no extra functionality.

\subsection{Continuation Passing}

\textbf{Case Study: Advancing Pages in the \emph{Story screen}}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{story-anim}
\caption{The \emph{Story screen}.
\label{fig.story}
}
\end{figure}

@@

\begin{figure*}[t]
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryScreenComponent::StoryScreenComponent (<...>) :
    <...>
{
    pages        = <...>; // vector with loaded pages
    current_page = pages.back(); // first loaded page
    displayed    = false; // if current is complete
    <...>
}

@@

void StoryScreenComponent::update (<...>) {
    <...>
    if (<all-words-appearing>) {
        displayed = true;
    }
}

@@

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Advancing pages in the \emph{Story Screen}.
\label{lst.story}
}
\end{figure*}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{story}
\caption{State machine for advancing pages in the \emph{Story screen}.
\label{fig.story}
}
\end{figure}

@@

\subsection{Dispatching Hierarchies}

\textbf{Case Study: Bomber `draw` and `update` callbacks}

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
class Bomber : public PinguAction {
    <...>
    Sprite sprite;
}

@@

void Bomber::draw (SceneContext& gc) {
    <...>
    gc.color().draw(sprite, <...>);
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName do
    <...>
    var&? Sprite sprite = spawn Sprite(<...>);
    <...>
end

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{hierarchy}
\caption{Dispatching chain for \code{update}.
\label{fig.hier}
}
\end{figure}

@@

\subsection{Lifespan Hierarchies}

\textbf{Case Study: The Pingus Container}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pingus_create_die-anim}
\caption{Creation and death of pingus.
\label{fig.pingus_create_die}
}
\end{figure}

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude
(Figure~\ref{pingus_create_die}).

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Pingu* PinguHolder::create_pingu (<...>) {
    <...>
    Pingu* pingu = new Pingu (<...>);
    pingus.push_back(pingu);
    <...>
}

void PinguHolder::update() {
    <...>
    while(pingu != pingus.end()) {
        (*pingu)->update();
        if ((*pingu)->get_status() == PS_DEAD) {
            pingu = pingus.erase(pingu);
        }
        <...>
        ++pingu;
    }
}

.
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Game (void) do
    <...>
    pool[] Pingu pingus;
    code/await Pingu_Spawn (<...>) do
        <...>
        spawn Pingu(<...>) in pingus;
    end
    <...>   // code invoking Pingu_Spawn
end

code/await Pingu (<...>) do
    <...>
    loop do
        await game.dt;
        if Pingu_Is_Out_Of_Screen() then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

In C++, the class \code{PinguHolder} is a container that holds all pingus
alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{games.patters} and
also occurs in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{pool}
\caption{Lifespan of dynamic instances.
\label{fig.pool}
}
\end{figure}

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen spawns a new \code{Pingu} on every invocation of
\code{Pingu\_Spawn}.
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its *natural termination*, the instance is automatically removed
from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

\subsection{Signaling Mechanisms}

\textbf{Case Study: Global Keys and the Options Menu}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{options-anim-opt}
\caption{The \emph{Mouse Grab} configuration option.
\label{fig.options}
}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{events}
\caption{Mutual dependency between signals.
\label{fig.events}
}
\end{figure}

@@

\begin{figure*}[t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
void GlbEvt::on_button_press (<...>) {
    <...>
    switch (event.keysym.sym) {
        case K_g:
            if (keys[K_LCTRL] || keys[K_RCTRL]) {
                cfgmgr.set_grab(
                    !cfgmgr.get_grab());
            }
            break;
        <...>
    }
}

@@

boost::signal<void(bool)> on_grab_change;

void CfgMgr::set_grab (bool v) {
    <...>
    if (v != get_grab()) {
        <...>   // the actual "grab" effect
        on_grab_change(v);
    }
}

@@

boost::signal<void (bool)> on_change;

void ChkBox::set (bool on, bool sndsig) {
    <...>   // switches the check box state
    if (sndsig) {
        on_change(on);
    }
}
